Computing likelihoods under the SMC for a general class of ARGs.

### 2024-04-03 paper.tex
Updates from Overleaf
review figures and likelihood section

In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineage with \emph{overlapping} genomic intervals may coalesce.
Surprisingly, although this is defined in terms of a backwards-in-time Markov process,
the resulting tree-valued process along the genome is also Markov.
Although \citet{mcvean_approximating_2005} verbally described the SMC 
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
(e.g., \comment{cite}).

@@

Coalescence occurs between
two overlapping lineages at rate $\lambda = 1/(2N_e)$
(if we assume a randomly mating, diploid population of constant size $N_e$).
The newly formed lineage acquires the union of both ordered intervals, % L is updated accordingly
and the original lineages are removed.
Although coalescence is reciprocal, here we'll define a strict total order on lineages
(see \ref{par:liks}) on all lineages in $L(t)$ based on their leftmost starting points.
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{ij}$,
where
\begin{equation} \label{def:coal}
I_{ij} = \begin{cases}
    1 & X_i > X_j \text{ and } s(X_i) \cap s(X_j) \neq \emptyset \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $j$ can coalesce \emph{into} lineaege $i$ if $I_{ij} = 1$
(and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).

@@

Finally, there is one more operation that allows the process to terminate:
portions of the edges of lineages that have entirely coalesced
(i.e., that are ancestral to all samples) are removed.
Formally, after any coalescence or recombination event,
the segment $[x,y)$ carried by the lineage
is replaced by the largest segment whose endpoints have not coalesced,
i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
and $y$ is replaced by $\sup\{ z \le y : N(t,z) > 1\}$.
\comment{explain the cases in which this happens: coalescence on one end,
and recobmbination in the middle of a trapped ancestral segment}

% note on the disjoint set of intervals for each lineage under the SMC
\comment{I'm not sure we need this here now?}
Note that 
an alternative definition could instead remove any sub-segment of a lineage
that has entirely coalesced.
This would produce a process in which each lineage is a collection
of disjoint intervals (instead of a single one).
As noted by \cite{mcvean_approximating_2005},
this implies that the backwards-in-time description of the
SMC is slightly different from the left-to-right description,
which would usually not include these segments.
This is depicted in fig.\ref{fig:smc-unary}:
for instance, node 8 is ancestral to all samples on the middle two trees.

@@

\comment{clarify what's different to McVean \& Cardin: unary nodes, maybe? and different to usual SMC stuff more generally}

@@

The process of recombination only affects one lineage at a time
(and does not even depend on the state of other lineages).
On the other hand, coalescence events necessarily involve more than one lineage.
However, the asymmetric way we define coalescence above
allows us to uniquely associate each event with a single lineage,
which allows us to cleanly factor the likelihood across edges and nodes.
The data structure we will compute the likelihood of
is a collection of \emph{edges} $E=\{e_i\}$,
where each edge is of the form $e_i = (p_i,c_i,x_i,y_i)$.
Recall that this means that node $c_i$ inherited genetic material from node $p_i$
over a region with span from $x_i$ to $y_i$ along the genome
(see Figure~\ref{fig:likelihood}).
So, $x_i$ and $y_i$ are genomic coordinates,
while $c_i$ and $p_i$ are in $N$, the set of \emph{nodes}
(i.e., distinct sampled or ancestral haplotypes).

We have defined the SMC process in terms of rates of various types of events.
In general, the likelihood for a continuous-time Markov process specified in this way
is $\exp(-\Lambda) \prod_i \lambda_i$,
where $\lambda_i$ is the rate of the $i^\text{th}$ realized event,
and $\Lambda$ is the sum of all rates of all possible events
integrated over the entire process
(also called the ``total hazard'').

The contributions to the likelihood from coalescent and recombination events
is straightforward, since each possible coalescence is associated with rate $\lambda$,
and each recombination with rate $r$.
Under the SMC, we only ever have two lineages coalescing at a time;
however, inferred ARGs may have more-than-binary mergers.
We treat these as if there were zero-length edges
between a sequence of mergers;
the order of these unresolved mergers does not affect the likelihood.
(Under the SMC the likelihood of a zero-length edge is zero;
but this is not special: since edge lengths are continuously distributed,
the likelihood of any particular value is also zero.)
Let $\deg_o(n) = \#\{e_i \in E : p_i=n\}$ be the out-degree of a node $n$,
i.e., the number of edges having $n$ as a parent.
Since mergers are binary in the SMC,
this must have been the result of $\deg(n) - 1$ coalescences,
and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
for nodes with any children.
Similarly,
let $\deg_i(n) = \#\{e_i \in E : c_i=n\}$ be the in-degree of a node $n$,
i.e., the number of edges having $n$ as a child.
Each additional edge that is ancestral to a given node
implies one additional ancestral recombination event,
and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood,
if the node has any parents.
Put together, these two contribute
\begin{equation}\label{eq:coal}
    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
    =
    r^{|E|-|N|+n_r} 
    \lambda^{|E|-|N|+n_s} 
\end{equation}
to the likelihood,
where $n_r$ is the number of nodes with in-degree 0 (roots),
and $n_s$ is the number of nodes with out-degree 0 (usually, samples).
(The simple form follows because each edge contributes exactly one
to some node's out-degree and some node's in-degree.)

@@

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$.
Under the SMC,
we therefore know there was no recombination between $x$ and $y$
on any of the lineages 
over this time span,
the probability of which is
$\exp(-r \times A)$, where $r$ is the recombination rate
and $A$ is the total area (span times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(\theta) = \exp(-r (y-x)(t_p - t_{c})) .
\end{equation}

Finally, we turn to the hazard associated with coalescence.
To do this, we need only consider what happens along the left side
of each edge.
The instantaneous rate of coalescence of a lineage whose left endpoint is at $x$
at a given time is equal to $\lambda$ multiplied
by the number of earlier lineages that overlap $x$ at that time.
Recall that we define a total order on edges,
by ordering by left endpoint and breaking ties by child
(i.e., $(p,c,x,y) < (p',c',x',y')$
if $x<x'$, or if $x=x'$ and $c<c'$).
Using this total order, we can define
$I_e(t)$ to be the total number of earlier edges that overlap edge $e$
and are present at time $t$.
Then, the cumulative hazard for coalescence of edge $e$
between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(u) du$.
If the edge $e = (p,c,x,y)$
is the leftmost segment of material inherited by node $c$
(e.g., the lefthand segment ancestral to node $c$
in Figure~\ref{fig:likelihood}),
then this edge simply contributes $\exp(-F(e,t_c,t_p))$ to the likelihood.
However, if the edge is \emph{not} the leftmost segment,
then it was initiated by a new recombination breakpoint,
and we need to integrate over possible times the breakpoint occurred.
If this is the case, there is another edge $(p',c,x',x)$ with the same child node,
immediately adjacent to the focal edge $(p,c,x,y)$.
The recombination event that split the two must have happened between $t_c$
and the smaller of $t_p$ and $t_{p'}$;
taking all this into consideration, the contribution to the likelihood here is
\begin{equation}\label{eq:depth}
B_e(\theta) = \begin{cases}
    e^{-\lambda F(e, t_c, t_p)}
        & \qquad \text{if } x=x_{c} \\
    \int_{t_c}^{t_{p} \wedge t_{p'}} e^{-\lambda F(e, s, t_{p})} ds
        & \qquad \text{if } x>x_{c} ,
\end{cases}
\end{equation}
where $x_c$ is the leftmost position of the edges whose children are $c$.

@@

\comment{This is the old equation; I think the $e^{-rs}$ should not be there,
    since it's already taken into account by \eqref{eq:span}.
    Plus, the $\lambda$ and $r$ factors are now elsewhere.}
\begin{equation}\label{eq:depth_old}
B_{(c, p)}(\theta) = \begin{cases}
e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c_s, s, t_{p})} ds \lambda^{I_{c_{s}d}(t_p)} & x=x_{c_{s}0}>x_{c0} \\
\end{cases}
\end{equation}
\comment{Do we still need this bit?}
Here, the (second) exponential term gives the probability of not observing a
coalescence event before $t_p$. In case of a recombination, $re^{-rs}$ further
gives the probability of observing a recombination
at time $s$ given a Poisson process along position $x$ with rate $r$.
The time of the recombination is integrated out.
The last term gives the point probability density of the coalescence event
that terminates the edge. The indicator function is used to resolve
the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.

The likelihood of ARG defined by the set of edges $E$ and nodes $N$
given parameters $\theta$ then is
\begin{equation}\label{eq:full-lik}
    \mathcal{L}(\mathcal{G}|\theta)
    =
    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(c, p) \in E} A_e(\theta) B_e(\theta) .
\end{equation}
Although the first two factors are easily expressed in terms of numbers of edges and nodes,
we will see below
that in practice it is convenient to absorb these terms into per-edge contributions.

\comment{TO DO: is this still true?}
Because the likelihood is computed based on edges,
and edges are no longer recorded once the local most recent common ancestor
has been reached, we will underestimate the number of positions where
a recombination could have occurred in case a recombination hits a
lineage with trapped non-ancestral material (see end \ref{par:description}).
If we assume that only a single recombination could have
occurred, then it suffices however to provide a correction factor $g$ in equation
\ref{eq:depth}. Here $g$ represents the
length of non-ancestral material along which a recombination would have resulted
in the same observed local sequence of tree changes.

Note that we are only considering the topology and branch length information of any ARG.
Extending this expression by adding the likelihood of observing a set of mutations
given an ARG is straightforward.
\comment{Perhaps give the expression, for completeness?}

\subsection{Algorithm} \label{par:algo}

@@

\comment{This is \emph{very} nice!}
This allows us to compute a likelihood in a single pass over all edges.
More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
the number of extant lineages in each internode interval (see fig \ref{fig:algo}).
When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along the internode intervals
spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
For each subsequently inserted edge, we first compute the likelihood contribution
of that edge by computing the integral
in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
\comment{At first I thought we didn't need this next sentence now, but maybe
assigning-$\lambda$-and-$r$ to each edge makes it easier to recompute the likelihood
for slices?}
We further keep track of
the last parent a child node was connected to to detect recombination events.

Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
through time by intersecting the internode intervals with the rate change intervals.
Rate changes along the genome can be incorporated similarly (see \ref{sec:slicing}).

@@

\subsection{Slicing the ARG} \label{sec:slicing}

% slicing of the ARG
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome. This tool therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution and context to a dataset of genetic variation.
It also opens up the opportunity for more efficient optimization
strategies during inference given that each slice of the ARG can be
processed independently.

From a data structure perspective, we can compute a likelihood
under the SMC for any valid tree sequence.
The \tskit library provides a wide range
of subsetting operations on tree sequences allowing us to slice
the ARG in time and/or space, as well as to reduce the number of
tracked samples. Any such subset of the tree sequence is itself
a valid tree sequence for which we can again compute a likelihood.
\comment{This next remark is totally not obvious, and a really nice observation;
maybe it needs a bit more explanation?}
\comment{Oh, I think it's not \emph{quite} true: if we slide at position $x$,
then any recombinations that happened at $x$ are now invisible, right?
So maybe you need to slice at non-integer positions, or do some bookeeping
about the number of edges with starts exactly at $x$?}
Remarkably, if we slice an ARG into disjoint pieces,
the likelihood of the full ARG
is the product of the likelihoods of each of the pieces.

Figure \ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate,
despite only using 10kb of sequence.

Although we can associate a likelihood with any valid tree sequences,
there are some caveats. By treating each edge independently we can
naturally deal with the various degrees of completeness to which inheritance
is documented by the different ARG inference methods: e.g., polytomies, absence
of locally unary coalescence nodes, unpreserved identify of internal nodes
across marginal trees, etcetera.
However, many of these imprecisions violate the implicit
assumptions of the likelihood computation.
\comment{Do we want these remarks in here, or as part of the Discussion?}
For example, the described algorithm assumes the
presence of ``locally unary'' nodes to correctly identify the
full extent of all lineages going back through time and therefore recombination events.
Without unary nodes though, our algorithm will detect a recombination
event along both coalescing lineages in case of a (sub)tree height changes along
the ARG as both edges would switch parent nodes. This can be mitigated by limiting
the number of 'inferrable' recombination events per coalescence event to 1
(see fig \ref{sup:fig:rec-correction}).
\comment{I'm not editing this in since I'm not sure I've got it right:
I think that as written, our algorithm would count an extra recombination
and and extra coalescence for any edge that got split up by removing the unary bits.
This could be corrected for in the algorithm, but it would require extra bookkeeping,
and make the likelihood not so easily a product over edges. (?)}

\comment{this was previously part of the discussion, have grouped it here.}
Although their presence is crucial to our ability to embed any local tree in the
ARG, the concept and their importance are still relatively new (but see
\citet{wong_general_2023}).
We have briefly touched upon the observation that in
the presence of unary nodes we seem to augment the left-to-right logic of the SMC
with some additional information. Yet, we believe that the topic of locally
unary nodes in general requires further exploration.
And more specifically, we see it as future work to fully quantify to what
extent the absence of these locally unary nodes will affect the shape of the
likelihood surface given any set of parameters.

%We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information and computed the corresponding likelihood using \texttt{msprime} \citep{baumdicker_efficient_2021}. After removing the recombination nodes the likelihoods as defined here were computed both with and without the unary nodes. Figure \ref{sup:fig:vs-hudson} shows that the proportion of variation explained is higher when the unary nodes are retained.

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
\caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
(5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples,
$r=2.5e-8$, $148$ edges in ARG). A single $N_e$ value was estimated for each
slice of the ARG as defined by the true population size change time points.
The inferred posterior distribution (right) using the SMC-likelihood was
obtained using a uniform prior. Number of edges per time slice
(from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
omitted for clarity.}
\label{fig:3-arg-slices}
\end{figure}

@@

Coalescence occurs between
two lineages with overlapping spans at rate $\lambda = 1/(2N_e)$
(if we assume a randomly mating, diploid population of constant size $N_e$).
The newly formed lineage acquires the union of both ordered intervals, % L is updated accordingly
and the original lineages are removed.
Although coalescence is reciprocal, here we'll define a strict total order on lineages
(see \ref{par:liks}) on all lineages in $L(t)$ based on their leftmost starting points.
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{ij}$,
where
\begin{equation} \label{def:coal}
I_{ij} = \begin{cases}
    1 & X_i > X_j \text{ and } s(X_i) \cap s(X_j) \neq \emptyset \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $j$ can coalesce \emph{into} lineaege $i$ if $I_{ij} = 1$
(and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).

>>>>

In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineage with \emph{overlapping} genomic intervals may coalesce.
Surprisingly, although this is defined in terms of a backwards-in-time Markov process,
the resulting tree-valued process along the genome is also Markov.
Although \citet{mcvean_approximating_2005} verbally described the SMC 
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend 
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.

@@

Coalescence occurs between
two overlapping lineages at rate $\lambda = 1/(2N_e)$
(if we assume a randomly mating, diploid population of constant size $N_e$).
The newly formed lineage acquires the union of both ordered intervals, % L is updated accordingly
and the original lineages are removed.
Although coalescence is reciprocal, here we'll define a strict total order
on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{ij}$,
where
\begin{equation} \label{def:coal}
I_{ij} = \begin{cases}
    1 & X_i > X_j \text{ and } s(X_i) \cap s(X_j) \neq \emptyset \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $j$ can coalesce \emph{into} lineage $i$ if $I_{ij} = 1$
(and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).

@@

Finally, there is one more operation that allows the process to terminate:
portions of the edges of lineages that have entirely coalesced
(i.e., that are ancestral to all samples) are removed.

Formally, after any coalescence or recombination event,
the segment $[x,y)$ carried by the lineage
is replaced by the largest (pair of) segment(s) whose endpoints have not
coalesced, i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
and $y$ is replaced by $\sup\{ z \le y : N(t,z) > 1\}$.
By reasoning about the process in this way we can keep representing 
all lineages by means of a single half-open interval throughout.

\comment{Adapt figure to represent this}

@@

@@

The process of recombination only affects one lineage at a time
(and does not even depend on the state of other lineages).
On the other hand, coalescence events necessarily involve more than one lineage.
However, the asymmetric way we define coalescence above
allows us to uniquely associate each event with a single lineage,
which allows us to cleanly factor the likelihood across edges and nodes.
The data structure we will compute the likelihood of
is a collection of \emph{edges} $E$,
where each edge $e$ is of the form $e = (p,c,x,y)$.
Recall that this means that node $c$ inherited genetic material from node $p$
over a region with span from $x$ to $y$ along the genome
(see Figure~\ref{fig:likelihood}).
So, $x$ and $y$ are genomic coordinates,
while $c$ and $p$ are in $N$, the set of \emph{nodes}
(i.e., distinct sampled or ancestral haplotypes).

We have defined the SMC process in terms of rates of various types of events.
In general, the likelihood for a continuous-time Markov process specified in this way
is $\exp(-\Lambda) \prod_i \lambda_i$,
where $\lambda_i$ is the rate of the $i^\text{th}$ realized event,
and $\Lambda$ is the sum of all rates of all possible events
integrated over the entire process
(also called the ``total hazard''). \comment{if we want to spell it out even further:
Note that the time to the next event is distributed as $\Lambda \exp(-\Lambda)$. And
the probablility that the next event is of type $i$ is $\lambda_i / \Lambda$.}

The contributions to the likelihood from coalescent and recombination events
is straightforward, since each possible coalescence is associated with rate $\lambda$,
and each recombination with rate $r$.
Under the SMC, we only ever have two lineages coalescing at a time;
however, inferred ARGs may have more-than-binary mergers.
We treat these as if there were zero-length edges
between a sequence of mergers;
the order of these unresolved mergers does not affect the likelihood.
(Under the SMC the likelihood of a zero-length edge is zero;
but this is not special: since edge lengths are continuously distributed,
the likelihood of any particular value is also zero.)
Let $\deg_o(n) = \#\{e \in E : p_e=n\}$ be the out-degree of a node $n$,
i.e., the number of edges having $n$ as a parent.
Since mergers are binary in the SMC,
this must have been the result of $\deg(n) - 1$ coalescences,
and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
for nodes with any children.
Similarly,
let $\deg_i(n) = \#\{e \in E : c_e=n\}$ be the in-degree of a node $n$,
i.e., the number of edges having $n$ as a child.
Each additional edge that is ancestral to a given node
implies one additional ancestral recombination event,
and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood,
if the node has any parents.
Put together, these two contribute
\begin{equation}\label{eq:coal}
    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
    =
    r^{|E|-|N|+n_r} 
    \lambda^{|E|-|N|+n_s} 
\end{equation}
to the likelihood,
where $n_r$ is the number of nodes with in-degree 0 (roots),
and $n_s$ is the number of nodes with out-degree 0 (usually, samples).
(The simple form follows because each edge contributes exactly one
to some node's out-degree and some node's in-degree.)

@@

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$.
Under the SMC,
we therefore know there was no recombination between $x$ and $y$
on any of the lineages 
over this time span,
the probability of which is
$\exp(-r \times A)$, where $r$ is the recombination rate
and $A$ is the total area (eligbile links times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
\end{equation}
Note that if a recombination did occurr, then the number of links not
affected by recombination is 1 less (see \ref{eq:depth}).

Finally, we turn to the hazard associated with coalescence.
To do this, we need only consider what happens along the left side
of each edge.
The instantaneous rate of coalescence of a lineage whose left endpoint is at $x$
at a given time is equal to $\lambda$ multiplied
by the number of earlier lineages that overlap $x$ at that time.
Recall that we define a total order on edges,
by ordering by left endpoint and breaking ties by child
(i.e., $(p,c,x,y) < (p',c',x',y')$
if $x<x'$, or if $x=x'$ and $c<c'$).
Using this total order, we can define
$I_e(t)$ to be the total number of earlier edges that overlap edge $e$
and are present at time $t$.
Then, the cumulative hazard for coalescence of edge $e$
between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(u) du$.
If the edge $e = (p,c,x,y)$
is the leftmost segment of material inherited by node $c$
(e.g., the lefthand segment ancestral to node $c$
in Figure~\ref{fig:likelihood}),
then no recombination occurred along this edge and 
this edge simply contributes $\exp(-F(e,t_c,t_p))$ to the likelihood.
However, if the edge is \emph{not} the leftmost segment,
then it was initiated by a new recombination breakpoint,
and we need to integrate over possible times the breakpoint occurred.
If this is the case, there is another edge $(p',c,x',x)$ with the same child node,
immediately adjacent to the focal edge $(p,c,x,y)$.
The recombination event that split the two must have happened between $t_c$
and the smaller of $t_p$ and $t_{p'}$. Note that two events must have happened
between $t_c$ and $\min(t_p, t_{p'})$; taking all this into consideration,
the contribution to the likelihood here is
\begin{equation}\label{eq:depth}
B_e(r, \lambda) = \begin{cases}
    e^{-\lambda F(e, t_c, t_p)}
        & \qquad \text{if } x=x_{c} \\
    \int_{t_c}^{t_{p} \wedge t_{p'}} e^{- r s} e^{-\lambda F(e, s, t_{p})} ds
        & \qquad \text{if } x>x_{c} ,
\end{cases}
\end{equation}
where $x_c$ is the leftmost position of the edges whose children are $c$.

@@

%\comment{This is the old equation; I think the $e^{-rs}$ should not be there,
%    since it's already taken into account by \eqref{eq:span}.
%    Plus, the $\lambda$ and $r$ factors are now elsewhere.}
%\begin{equation}\label{eq:depth_old}
%B_{(c, p)}(\theta) = \begin{cases}
%e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
%\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c_s, s, t_{p})} ds \lambda^{I_{c_{s}d}(t_p)} & x=x_{c_{s}0}>x_{c0} \\
%\end{cases}
%\end{equation}
%\comment{Do we still need this bit?}
%Here, the (second) exponential term gives the probability of not observing a
%coalescence event before $t_p$. In case of a recombination, $re^{-rs}$ further
%gives the probability of observing a recombination
%at time $s$ given a Poisson process along position $x$ with rate $r$.
%The time of the recombination is integrated out.
%The last term gives the point probability density of the coalescence event
%that terminates the edge. The indicator function is used to resolve
%the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.

The likelihood of ARG defined by the set of edges $E$ and nodes $N$
given parameters $\theta$ then is
\begin{equation}\label{eq:full-lik}
    \mathcal{L}(\mathcal{G}|r, \lambda)
    =
    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(c, p) \in E} A_e(r) B_e(r, \lambda) .
\end{equation}
Although the first two factors are easily expressed in terms of numbers of edges and nodes,
we will see below
that in practice it is convenient to absorb these terms into per-edge contributions.

Note that we are only considering the topology and branch length information of any ARG.
Extending this expression by adding the likelihood of observing a set of mutations
given an ARG is given by
\begin{equation}\label{eq:lik-mut}
    \mathcal{L}(\mathcal{D}|\mathcal{G}, \mu)
    =
    \frac{1}{\mathcal{M}!} \prod_{(c, p) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}} .
\end{equation}
where $\mathcal{M}$ the total number of mutations and $m_e$
the number of mutations along edge $(c,p)$ \citep{mahmoudi_bayesian_2022}.

% % % % % % % % % % % % % % % % % %
\subsection{Algorithm} \label{par:algo}

@@

This allows us to compute a likelihood in a single pass over all edges.
More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
the number of extant lineages in each internode interval (see fig \ref{fig:algo}).
When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along the internode intervals
spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
For each subsequently inserted edge, we first compute the likelihood contribution
of that edge by computing the integral
in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
We further keep track of
the last parent a child node was connected to to detect recombination events.
Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
through time by intersecting the internode intervals with the rate change intervals.
Rate changes along the genome can be incorporated similarly (see \ref{sec:slicing}).

@@

% % % % % % % % % % % % % % % % % %
\subsection{Slicing the ARG} \label{sec:slicing}

% slicing of the ARG
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome.
In the case of a fully precise ARG, with a known time to recombination,
this is true as long as the slicing operation does not erase the
information on whether an edge is associated with the lefthand or
righthand segment generated by a recombination event. In the absence
of the exact time to recombination, the integral in
eq.\ \ref{eq:depth} means that the likelihood
for the entire ARG does not factorize into the likelihoods of these
individual slices. More precisely, the likelihood associated with any
edge that is initiated by a recombination breakpoint cannot be factorized
into the product of two integrals of the same form.

However, as long as the individual slices of the ARG contain sufficient
information we can approximate the true integral as the product of the
integral of the slices. Figure \ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate, despite only using 10kb of sequence.
This approach therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution and context to a dataset of genetic variation.
It also opens up the opportunity for more efficient optimization
strategies during inference given that each slice of the ARG can be
processed independently. And it further highlights the flexibility
of the algorithm. Simply stated and from a data structure perspective,
we can compute a likelihood under the SMC for any valid tree sequence.
The \texttt{tskit} library provides a wide range
of subsetting operations on tree sequences allowing us to slice
the ARG in time and/or space, as well as to reduce the number of
tracked samples. Any such subset of the tree sequence is itself
a valid tree sequence for which we can again compute a likelihood.

% Although we can associate a likelihood with any valid tree sequences,
% there are some caveats. By treating each edge independently we can
% naturally deal with the various degrees of completeness to which inheritance
% is documented by the different ARG inference methods: e.g., polytomies, absence
% of locally unary coalescence nodes, unpreserved identify of internal nodes
% across marginal trees, etcetera.
% However, many of these imprecisions violate the implicit
% assumptions of the likelihood computation.
% \comment{Do we want these remarks in here, or as part of the Discussion?}
% 
% For example, the described algorithm assumes the
% presence of ``locally unary'' nodes to correctly identify the
% full extent of all lineages going back through time and therefore recombination events.
% Without unary nodes though, our algorithm will detect a recombination
% event along both coalescing lineages in case of a (sub)tree height changes along
% the ARG as both edges would switch parent nodes. This can be mitigated by limiting
% the number of 'inferrable' recombination events per coalescence event to 1
% (see fig \ref{sup:fig:rec-correction}).
% \comment{I'm not editing this in since I'm not sure I've got it right:
% I think that as written, our algorithm would count an extra recombination
% and and extra coalescence for any edge that got split up by removing the unary bits.
% This could be corrected for in the algorithm, but it would require extra bookkeeping,
% and make the likelihood not so easily a product over edges. (?)}
% 
% \comment{this was previously part of the discussion, have grouped it here.}
% Although their presence is crucial to our ability to embed any local tree in the
% ARG, the concept and their importance are still relatively new (but see
% \citet{wong_general_2023}).
% We have briefly touched upon the observation that in
% the presence of unary nodes we seem to augment the left-to-right logic of the SMC
% with some additional information. Yet, we believe that the topic of locally
% unary nodes in general requires further exploration.
% And more specifically, we see it as future work to fully quantify to what
% extent the absence of these locally unary nodes will affect the shape of the
% likelihood surface given any set of parameters.

%%% remarks on precision of information + useage of likelihood to either
%%% estimate parameters (not goal here) or use as instrument to improve accuracy to infer ARGs

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
    \caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
    for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
    (5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples,
    $r=2.5e-8$, $148$ edges in ARG. A single $N_e$ value was estimated for each
    slice of the ARG as defined by the true population size change time points.
    The inferred posterior distribution (right) using the SMC-likelihood was
    obtained using a uniform prior. Number of edges per time slice
    (from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
    omitted for clarity.}
    \label{fig:3-arg-slices}
\end{figure}

\section{Likelihood surface for larger class of ARGs}

% flexibility
% presence of polytomies
% lack of unary nodes

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{figures/likelihood_surface.png}
    \caption{Simulated ARG under SMC ($L=1$Mb, $100$ diploid samples,
    $r=2.5e-8$, $N_e=1e4$). Likelihood-ratio contour plot. Currently two contours,
    with and without locally unary nodes.}
    \label{fig:lik-surface}
\end{figure}

@@

Coalescence occurs between
two lineages with overlapping spans at rate $\lambda = 1/(2N_e)$
(if we assume a randomly mating, diploid population of constant size $N_e$).
The newly formed lineage acquires the union of both ordered intervals, % L is updated accordingly
and the original lineages are removed.
Although coalescence is reciprocal, here we'll define a strict total order
on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{ij}$,
where
\begin{equation}
I_{ij} = \begin{cases}
    1 & X_i > X_j \text{ and } s(X_i) \cap s(X_j) \neq \emptyset \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $j$ can coalesce \emph{into} lineage $i$ if $I_{ij} = 1$
(and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).
